Graphical Tests for Power Comparison of Competing Designs

### 2012-06-27 review/response.txt

I would like to thank the reviewers for the great discussion of the paper, and a lot of helpful suggestions!

We have made the following changes:

All of the main suggestions are implemented:
 - in the presentation of figure 8, differences between power results are now presented using a Benjamini-Hochberg adjustment for the false discovery rate.
   Since BH adjustments are based on individual p-values (whereas the Bonferroni adjustment results in a global scaling of the intervals), it was necessary to change the presentation in the figure slightly - we now show the significant differences using a letter-based approach as suggested by H.-P. Piepho (2004).

 - the paper is re-organized and the two studies are now presented one after the other.

 - the discussion regarding pie vs barcharts is removed - we'd like to thank the reviewers for pointing out the additional references.

Changes based on suggestions of individual reviewers:
 - the blue scale for the pictures of the wind direction was darkened a bit, which makes the small levels better distinguishable both in print and pdf.

 - reporting the log: we indeed used log base 10 for figure 6 - we changed this picture to natural log for consistency.

 - Euclidean vs Cartesian coordinates - agreed, and changed

 - language concerning null hypothesis and alternative in 3.2 is fixed.

 - The scale for the self-reported confidence is included explicitly now.

 -  the design choices for the second study were based on designs commonly used; we have addressed this in the paper.

 - conclusion section is strengthened and also focuses on the results from the experiment rather than just their setup.

Discussion:

Q: Would it be more understandable and specific to characterize the power
  model as a hierarchical linear model with random intercept instead of a
  generalized linear model? If I understand correctly, the HLM with random
  intercept is specifically what they are using. If not, the authors can
  explain to the reviewers why I am wrong and not make the change.

A: What we are using in estimating power is actually a generalized linear model with a random intercept - since power has to be a value between 0 and 1, we are using a logistic regression instead of a linear model.

Q: The first study does not state clearly what constitutes "best" time for
  minutes between wheel events. You claim that it is clear that Seattle
  functions at its best with winds coming from the East but I'm not sure if
  that means that you just want the lowest number of minutes or if there is
  an optimal number. The hypotheses just address "differences" but not any
  specific time interval in general.

A: We were trying to avoid the phrase 'best' - as we are, indeed, just looking at efficiency ('its best' has been changed in the paper to 'most efficiently') - i.e. less time between wheel events is considered 'better'. (Note: Seattle has several runways, so a time of 0 between wheel events is not as unsafe as it might sound).

Hans-Peter Piepho (2004), An Algorithm for a Letter-Based Representation of All-Pairwise Comparisons, Journal of Computational and Graphical Statistics, 13(2), 456–466.

### 2012-06-27 review/comparison-review.tex
revision

%% Abstract section.
\abstract{
Lineups  \cite{buja:2009, wickham:2010} have been established as tools for visual testing
similar to standard statistical inference tests, allowing us to
evaluate the validity of graphical findings in an objective manner. In
simulation studies  \cite{majumder:2011} lineups have been shown as being efficient: the
power of visual tests is comparable to classical tests while being
much less stringent in terms of distributional assumptions made. This
makes lineups versatile, yet powerful, tools in situations where
conditions for regular statistical tests are not or cannot be met. In
this paper we introduce lineups as a tool for evaluating the power of
competing graphical designs. We highlight some of the theoretical
properties and then show results from two studies evaluating competing
designs: both studies are designed to go to the limits of our
perceptual abilities to highlight differences between designs. We use
both accuracy and speed of evaluation as measures of a successful
design. The first study compares the choice of coordinate system:
polar versus Euclidean coordinates. The results show strong support in
favor of Euclidean coordinates in finding fast and accurate answers to
spotting patterns. The second study is aimed at finding shift
differences between distributions. Both studies are motivated by data
problems that we have recently encountered, and explore using
simulated data to evaluate the plot designs under controlled
conditions.  Amazon Turk is used to conduct the studies.  The lineups
provide an effective mechanism for objectively evaluating plot
designs.
} % end of abstract

>>>>

%% Abstract section.
\abstract{
Lineups  \cite{buja:2009, wickham:2010} have been established as tools for visual testing
similar to standard statistical inference tests, allowing us to
evaluate the validity of graphical findings in an objective manner. In
simulation studies  \cite{majumder:2011} lineups have been shown as being efficient: the
power of visual tests is comparable to classical tests while being
much less stringent in terms of distributional assumptions made. This
makes lineups versatile, yet powerful, tools in situations where
conditions for regular statistical tests are not or cannot be met. In
this paper we introduce lineups as a tool for evaluating the power of
competing graphical designs. We highlight some of the theoretical
properties and then show results from two studies evaluating competing
designs: both studies are designed to go to the limits of our
perceptual abilities to highlight differences between designs. We use
both accuracy and speed of evaluation as measures of a successful
design. The first study compares the choice of coordinate system:
polar versus cartesian coordinates. The results show strong support in
favor of cartesian coordinates in finding fast and accurate answers to
spotting patterns. The second study is aimed at finding shift
differences between distributions. Both studies are motivated by data
problems that we have recently encountered, and explore using
simulated data to evaluate the plot designs under controlled
conditions.  Amazon Mechanical Turk (MTurk) is used to conduct the studies.  The lineups
provide an effective mechanism for objectively evaluating plot
designs.
} % end of abstract

### 2012-06-27 review/experiment-two.tex
revision

% !TEX root = comparison.tex
\subsection{Comparing Distribution Centers}

Since each participant provides results from nine lineups, we can use a generalized linear mixed effects model of power and account for individuals' abilities by including a subject-specific random intercept. We included all of the above described covariates  in the model, both as main effects and as two-way interaction with the design to assess their impact on the power of each of the designs.
Figure \ref{fig:power2} summarizes the modeling results as shown in table \ref{tbl:power2}. Generally we can see that the power of all four designs increases with an increase in effect size $d$ (on the $x$ axis), as well as an increasing sample size $n_1$ (top to bottom facetting). Generally, an increase of the second group also is beneficial to power. Unlike all of the other designs,  dotplots actually lose power significantly as the second group increases. 
The reason behind this phenomenon might be visual complexity: dotplots are the only design that show individuals. With an increase in group size, visual complexity increases with it, which could hamper the observer in evaluating shifts.

\begin{table}[ht]
\begin{center}
\resizebox{\linewidth}{!} {
\begin{tabular}{lrrrrrl}
  \hline
& & Estimate & Error & $t$-value & \multicolumn{2}{l}{approx $p$-value} \\   \hline
\\[-8pt]
\bf design& Intercept & -3.11 & 0.41 & -7.61 & 0.00 & ***\\ 
&  density & 0.05 & 0.57 & 0.09 & 0.93 \\ 
&  dotplot & 0.46 & 0.57 & 0.81 & 0.42\\ 
&  histogram & 0.06 & 0.58 & 0.11 & 0.91 \\ [2pt]
\multicolumn{2}{l}{\bf covariates}\\
&  d & 1.77 & 0.34 & 5.18 & 0.00 & ***\\ 
&  n1$_{100}$ & 1.84 & 0.19 & 9.73 & 0.00 & ***\\ 
&  ratio & 0.84 & 0.33 & 2.56 & 0.01 & **\\ 
\multicolumn{2}{l}{\bf interactions}\\
&  density:d & -0.60 & 0.47 & -1.27 & 0.20 \\ 
&  dotplot:d & 0.03 & 0.46 & 0.07 & 0.95  \\ 
&  histogram:d & -0.28 & 0.48 & -0.59 & 0.56 \\ [1pt]
&  density:n1$_{100}$ & -0.10 & 0.26 & -0.36 & 0.72\\ 
&  dotplot:n1$_{100}$ & -0.77 & 0.26 & -3.00 & 0.00 & ***\\ 
&  histogram:n1$_{100}$ & -0.69 & 0.26 & -2.64 & 0.01 & ** \\ [1pt]
&  density:ratio & -0.28 & 0.47 & -0.59 & 0.56 \\ 
&  dotplot:ratio & -1.01 & 0.46 & -2.19 & 0.03 & * \\ 
&  histogram:ratio & -0.28 & 0.47 & -0.59 & 0.55 \\ 
   \hline
\\[-5pt]
   \multicolumn{5}{l}{Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1}
\end{tabular}}
\end{center}
\caption{\label{tbl:power2}Results from a Generalized Linear Mixed Effects Model of power of a  lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size. 
Note that for all terms involving $n_1$ estimate and error are multiplied by a factor of 100. 
Reference group for all effects are boxplot designs. The model fit is based on 2513 lineup evaluations by 208 participants. }
\end{table}

The two highly summarized designs benefit the most from an increase in sample size: both density charts and boxplots profit significantly more than histograms and dotplots. This might, again, be a hint towards a trade-off between visual complexity and amount of information shown as the basis for making decisions.

@@

\begin{table}[ht]
\begin{center}
\resizebox{\linewidth}{!} {
\begin{tabular}{lrrrrrl}
  \hline
& & Estimate & Error & $t$-value & \multicolumn{2}{l}{approx $p$-value} \\   \hline
\\[-8pt]
\bf design& Intercept & 3.85 & 0.09 & 42.21 & 0.00 & ***\\ 
&  density & 0.01 & 0.12 & 0.07 & 0.94 \\ 
&  dotplot & 0.03 & 0.12 & 0.26 & 0.80 \\ 
&  histogram & 0.23 & 0.12 & 1.97 & 0.05 & . \\ [2pt]
\multicolumn{2}{l}{\bf covariates}\\
&      d & -0.23 & 0.08 & -3.02 & 0.00 & ***\\ [1pt]
&    n1$_{100}$ & 0.07 & 0.04 & 1.70 & 0.09 & .\\ [1pt]
&  ratio & -0.04 & 0.02 & -1.48 & 0.14  \\ [2pt]
\multicolumn{2}{l}{\bf interactions}\\
&  density:d & 0.11 & 0.11 & 1.02 & 0.31 \\ 
&  dotplot:d & 0.20 & 0.11 & 1.86 & 0.06 & .  \\ 
&  histogram:d & -0.01 & 0.11 & -0.09 & 0.93 \\ [1pt]
&  density:n1$_{100}$ & -0.12 & 0.06 & -2.11 & 0.03 & *\\ 
&  dotplot:n1$_{100}$ & -0.16 & 0.06 & -2.83 & 0.00 & ***\\ 
&  histogram:n1$_{100}$ & -0.05 & 0.06 & -0.93 & 0.35 \\ [1pt]
&  density:ratio & 0.04 & 0.03 & 1.18 & 0.24 \\ 
&  dotplot:ratio & 0.02 & 0.03 & 0.65 & 0.52 \\ 
&  histogram:ratio & 0.00 & 0.04 & 0.04 & 0.97 \\ 
   \hline
\\[-5pt]
   \multicolumn{5}{l}{Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1}
\end{tabular}}
\end{center}
\caption{\label{tbl:speed2}Results from a Linear Mixed Effects Model of (log) time to answer a lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size. Estimate and error for all terms involving $n_1$ are multiplied by a factor of 100 for comparability.
Reference group for all effects are boxplot designs. }
\end{table}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{conf-exp2} 
   \caption{Barcharts of marginal distributions of self-reported confidence by design. The distributions display prominent differences.}
   \label{fig:conf-margins}
\end{figure}

>>>>

% !TEX root = comparison-review.tex
\subsection{Study II: Comparing Distribution Centers}
\subsubsection{Study Setup}
In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in Figure \ref{fig:expii}, in assessing a mean shift between distributions.

\noindent The  hypotheses for each lineup are

\begin{tabular}{rl}
$H_o:$&{\em centers of the two groups are the same}, \\
$H_a:$&{\em center of the blue group is shifted to the right}.\\[3pt]
\end{tabular}
%\vspace{-0.1in}
%\vspace{-0.1in}

As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8. 1.0,$ and $1.2$, as well as the sample size.  Since we are dealing with two groups of points, we varied both the {\em size of the larger group}, $n_1 = 15, 45$, and 135, as well as the {\em relative size} of the second group ({\em ratio}), with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$. 
For each of these combinations, we created three replicate data sets with values from an exponential distribution with rate parameter $\lambda=1$ (for the first group) and rate parameter $\lambda_2=1/(d+1)$, which corresponds to a mean shift of $d$, for the second groups. 
Based on each data set we created four lineup charts, one for each of the competing designs. This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data. According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard). Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allow for data quality checks.
\begin{figure} [hbtp]
   \centering
   \includegraphics[width=0.45\linewidth]{hist-conc} 
   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
   \includegraphics[width=0.45 \linewidth]{density-conc} 
   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
\vspace{-0.1in}
   \caption{Overview of all four competing designs for displaying differences in distributions.}
   \label{fig:expii}
\end{figure}

All of the designs were chosen based on common usage. The main difference between the designs is their varying level of data summarization.

 From most to least summarized we consider these designs:

\begin{tabular}{lp{0.33\textwidth}}\\[-5pt]
\em boxplots: & five-point summary statistics (and a number of  outliers),\\
\em density plots: & need -- mathematically -- all of the data points, but  their smoothness and continuity makes them appear visually quite simple,\\
\end{tabular}
\begin{tabular}{lp{0.33\textwidth}}
 {\em histograms}: &  rendered by a frequency breakdown for a specified number of bins ($10 \times 2$, in the study),\\
 {\em dotplots}: & all $n$ data points necessary for rendering.\\[3pt]
\end{tabular}

\noindent For the boxplot and dotplot, the vertical axis is free to be used for another variable, {\em group}, so the data can be split apart. For the histogram and density the vertical axis is used for count or frequency, and is not available for group information.
Based on the assumption that charts are most effective \citep{Kosslyn:2006}, when they show enough information to solve the task at hand, but not too much to hamper the observer, we anticipated results to reflect the above order.

\subsubsection{Results}

Since each participant provides results from nine lineups, we can use a generalized linear mixed effects model of power and account for individuals' abilities by including a subject-specific random intercept. All of the above described covariates are considered  in the model, both as main effects and as two-way interaction with the design to assess their impact on the power of each of the designs.
Figure \ref{fig:power2} summarizes the modeling results as shown in table \ref{tbl:power2}. Generally we can see that the power of all four designs increases with an increase in effect size $d$ (on the $x$ axis), as well as an increasing sample size $n_1$ (top to bottom facetting). Generally, an increase of the second group also is beneficial to power. Dotplots are an exception to this -- they actually lose power significantly as the second group increases. This might be due to visual complexity, as discussed above.
%The reason behind this phenomenon might be visual complexity \citep{Kosslyn:2006}: dotplots are the only design that show individuals. With an increase in group size, visual complexity increases with it, which could hamper the observer in evaluating shifts.

\begin{table}[ht]
\begin{center}
\resizebox{\linewidth}{!} {
\begin{tabular}{lrrrrrl}
  \hline
& & Estimate & Error & $t$-value & \multicolumn{2}{l}{approx $p$-value} \\   \hline
\\[-8pt]
\bf design& Intercept & -3.11 & 0.41 & -7.61 & 0.00 & ***\\ 
&  density & 0.05 & 0.57 & 0.09 & 0.93 \\ 
&  dotplot & 0.46 & 0.57 & 0.81 & 0.42\\ 
&  histogram & 0.06 & 0.58 & 0.11 & 0.91 \\ [2pt]
\multicolumn{2}{l}{\bf covariates}\\
&  d & 1.77 & 0.34 & 5.18 & 0.00 & ***\\ 
&  n1$_{100}$ & 1.84 & 0.19 & 9.73 & 0.00 & ***\\ 
&  ratio & 0.84 & 0.33 & 2.56 & 0.01 & **\\ 
\multicolumn{2}{l}{\bf interactions}\\
&  density:d & -0.60 & 0.47 & -1.27 & 0.20 \\ 
&  dotplot:d & 0.03 & 0.46 & 0.07 & 0.95  \\ 
&  histogram:d & -0.28 & 0.48 & -0.59 & 0.56 \\ [1pt]
&  density:n1$_{100}$ & -0.10 & 0.26 & -0.36 & 0.72\\ 
&  dotplot:n1$_{100}$ & -0.77 & 0.26 & -3.00 & 0.00 & ***\\ 
&  histogram:n1$_{100}$ & -0.69 & 0.26 & -2.64 & 0.01 & ** \\ [1pt]
&  density:ratio & -0.28 & 0.47 & -0.59 & 0.56 \\ 
&  dotplot:ratio & -1.01 & 0.46 & -2.19 & 0.03 & * \\ 
&  histogram:ratio & -0.28 & 0.47 & -0.59 & 0.55 \\ 
   \hline
\\[-5pt]
   \multicolumn{5}{l}{Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1}
\end{tabular}}
\end{center}
\vspace{-0.2in}
\caption{\label{tbl:power2}Results from a Generalized Linear Mixed Effects Model of power of a  lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size. 
Note that for all terms involving $n_1$ estimate and error are multiplied by a factor of 100. 
Reference group for all effects are boxplot designs. The model fit is based on 2513 lineup evaluations by 208 participants. }
\end{table}

Visually, density plots seem to 
The two highly summarized designs benefit the most from an increase in sample size: both density charts and boxplots profit significantly more than histograms and dotplots. This might, again, be a hint towards a trade-off between visual complexity and amount of information shown as the basis for making decisions.

@@

\begin{table}[ht]
\begin{center}
\resizebox{\linewidth}{!} {
\begin{tabular}{lrrrrrl}
  \hline
& & Estimate & Error & $t$-value & \multicolumn{2}{l}{approx $p$-value} \\   \hline
\\[-8pt]
\bf design& Intercept & 3.85 & 0.09 & 42.21 & 0.00 & ***\\ 
&  density & 0.01 & 0.12 & 0.07 & 0.94 \\ 
&  dotplot & 0.03 & 0.12 & 0.26 & 0.80 \\ 
&  histogram & 0.23 & 0.12 & 1.97 & 0.05 & . \\ [2pt]
\multicolumn{2}{l}{\bf covariates}\\
&      d & -0.23 & 0.08 & -3.02 & 0.00 & ***\\ [1pt]
&    n1$_{100}$ & 0.07 & 0.04 & 1.70 & 0.09 & .\\ [1pt]
&  ratio & -0.04 & 0.02 & -1.48 & 0.14  \\ [2pt]
\multicolumn{2}{l}{\bf interactions}\\
&  density:d & 0.11 & 0.11 & 1.02 & 0.31 \\ 
&  dotplot:d & 0.20 & 0.11 & 1.86 & 0.06 & .  \\ 
&  histogram:d & -0.01 & 0.11 & -0.09 & 0.93 \\ [1pt]
&  density:n1$_{100}$ & -0.12 & 0.06 & -2.11 & 0.03 & *\\ 
&  dotplot:n1$_{100}$ & -0.16 & 0.06 & -2.83 & 0.00 & ***\\ 
&  histogram:n1$_{100}$ & -0.05 & 0.06 & -0.93 & 0.35 \\ [1pt]
&  density:ratio & 0.04 & 0.03 & 1.18 & 0.24 \\ 
&  dotplot:ratio & 0.02 & 0.03 & 0.65 & 0.52 \\ 
&  histogram:ratio & 0.00 & 0.04 & 0.04 & 0.97 \\ 
   \hline
\\[-5pt]
   \multicolumn{5}{l}{Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1}
\end{tabular}}
\end{center}
\vspace{-0.2in}
\caption{\label{tbl:speed2}Results from a Linear Mixed Effects Model of (log) time to answer a lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size. Estimate and error for all terms involving $n_1$ are multiplied by a factor of 100 for comparability.
Reference group for all effects are boxplot designs. }
\end{table}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.95\linewidth]{conf-exp2} 
\vspace{-0.25in}
   \caption{Barcharts of marginal distributions of self-reported confidence by design. The distributions display prominent differences.}
   \label{fig:conf-margins}
\end{figure}

@@

### 2012-06-27 review/intro.tex
revision

% !TEX root = comparison.tex
%% \section{Introduction} %for journal use above \firstsection{..} instead
%We introduce and discuss a tool for comparing different charts that are renderings of the same data in order to evaluate them for efficiency.
So, you've just spent countless hours burning the late night candle
wax and discovered several astonishing features in the data. Your
exploratory graphics are quick to produce, interactive, flexible, and
fabulous for discovery, but now you need to present the information to
your boss, the co-workers, a class of students, or get it published in
that top-rated journal. Those exploratory graphics are unlikely to
communicate the discovered information to your audience efficiently,
or elegantly.  To decide on the best way to present the specific
information there can be many types of plots to choose from. The
decision process is assisted by past experience, personal aesthetic
preferences, the characteristics of the audience, and a good working
knowledge of general perceptual strengths and weaknesses of particular
displays \cite{cleveland:1984,heer:2010}. Perceptual studies are
thinly spread for data visualization and will probably
only cover very broad perceptual principles, primarily because they
tend to be based solely on simulated data. In order to study
perceptual principles in a controlled experimental setting simulated
data is unavoidable. This provides principles that may not apply
closely enough to a specific analysis task.  Faced with design
decisions for a very specific task requires a closer level of
detail. Focus groups \cite{Mazza07} can find small problems with
designs and move us closer to a final handful of possible
designs. Usability studies and case studies
(e.g. \cite{Plaisant:2004}, \cite{Shneiderman:2006}) typically focus
on the use of particular software for solving data exploration tasks.
These approaches lack broad audience validity and
objectiveness. Several papers indicate continuing issues arising with,
and suggestions for improving, evaluation of information
visualizations
\cite{North:2006,Ellis:2006,Viegas:2006,Munzer:2009}.

@@

Here we suggest a new approach using recent research on visual inference \cite{buja:2009,wickham:2010}. In visual inference, data plots are considered to be test statistics, and these are compared with plots of data generated from a null hypothesis using a lineup. The null hypothesis underlying a lineup is that there is no real structure visible in the data plot, that any patterns seen are consistent with randomness, or from a known model. If the null hypothesis is true then the plot of the data will not be distinguishable form the plots of null data. A lineup is similar to the police lineup, and this is where the name comes from. The suspect (the plot of the data) is placed randomly in a field of plots of null data, and an impartial viewer is asked to identify the plot that is the most different. If the viewer identifies the suspect it lends statistical significance to the conclusion that the data is not consistent with the null data. Figure \ref{lineup} has an example of a lineup. Overlaid density plots are used to display the distribution of two groups (blue, red) of data. We are interested to know whether the data has a significant difference between the centers of the two groups. Plot numbered 13 contains the data. Did you pick that one? In this plot the blue group has a flat density whereas the density of red group is more peaked and concentrated to the left of the blue.

\begin{figure}[htbp]
   \centering
   \includegraphics[height=0.95\linewidth]{density_1_45_15_20_13.png} 
   \caption{Example lineup using density plots: one plot of data
     embedded with nineteen plots of null data. Which plot is
     most different from the others? (See the text for the answer.)}
   \label{lineup}
\end{figure}

The previous work \cite{buja:2009,wickham:2010} describes the use of lineups for evaluating discoveries made in exploratory analysis. Here we describe the use of lineups for evaluating plot design for communicating findings visually. Different lineups are created, using the same data, same null plots and positions within the lineup, but different plot designs. Comparing how long and how accurately the viewers take to report the feature of interest will assess which design is better for the task. To investigate the benefit of lineups for plot design evaluation we have conducted an experiment that also contains simulated data, where the data is constructed to match and expand on the features of interest in the data. For example, in a simple situation, where the discovered feature of interest is a difference between the centers of two distributions, simulated data would be constructed roughly matching the distribution of the data that varies the distance between centers, and also spread and sample size. Lineups of these simulated data (embedded with simulated null data plots) are generated and evaluated along with the lineups of data. The simulation study provides a backdrop for the real data, which enables the broad applicability to be studied and a gauge for the strength of the pattern in the data.

@@

Lineups provide a way to statistically quantify the significance of the finding \cite{majumder:2011}: think of the lineup as a set of $m$ plots, one of which is the data. (In Figure \ref{lineup}, $m=20$.)  The probability that an observer picks the data plot just by chance, i.e. when it is really *not* different from the other null plots, is $1/m$.  If an observer is able to identify the data plot from the lineup, we reject the null hypothesis.  This sets our Type I error rate, $\alpha$, at a level of $1/m$.

@@

The power of lineups is used in the work presented here to evaluate the effectiveness of different plot designs. This paper describes two examples where we had real data analysis problems and decisions to make in order to communicate the results. Section \ref{designs} describes the process for comparing designs. Section \ref{setup} describes the two data analysis problems and the experiments conducted to evaluate the plot designs. Section \ref{results} presents the findings of the experiments, and Section \ref{conclusions} presents conclusions and suggestions for future use.

@@

While this allows a direct comparison of the designs, we cannot adjust for the individuals' perceptual abilities.
In the case that we have multiple responses from each person (i.e. data is collected on several {\it different} lineup tasks), we can estimate for their perceptual ability and correct power differences between competing designs accordingly, e.g. by modeling power using  a subject-specific random intercept in a generalized linear model.

>>>>

% !TEX root = comparison-review.tex
%% \section{Introduction} %for journal use above \firstsection{..} instead
%We introduce and discuss a tool for comparing different charts that are renderings of the same data in order to evaluate them for efficiency.
So, you've just spent countless hours burning the late night candle
wax and discovered several astonishing features in the data. Your
exploratory graphics are quick to produce, interactive, flexible, and
fabulous for discovery, but now you need to present the information to
your boss, the co-workers, a class of students, or get it published in
that top-rated journal. Those exploratory graphics are unlikely to
communicate the discovered information to your audience efficiently,
or elegantly.  To decide on the best way to present the specific
information there can be many types of plots to choose from. The
decision process is assisted by past experience, personal aesthetic
preferences, the characteristics of the audience, and a good working
knowledge of general perceptual strengths and weaknesses of particular
displays \cite{cleveland:1984, simkin:1987,spence:1991, heer:2010}. Perceptual studies are
thinly spread for data visualization and will probably
only cover very broad perceptual principles, primarily because they
tend to be based solely on simulated data. In order to study
perceptual principles in a controlled experimental setting simulated
data is unavoidable. This provides principles that may not apply
closely enough to a specific analysis task.  Faced with design
decisions for a very specific task requires a closer level of
detail. Focus groups \cite{Mazza07} can find small problems with
designs and move us closer to a final handful of possible
designs. Usability studies and case studies
(e.g. \cite{Plaisant:2004}, \cite{Shneiderman:2006}) typically focus
on the use of particular software for solving data exploration tasks.
These approaches lack broad audience validity and
objectiveness. Several papers indicate continuing issues arising with,
and suggestions for improving, evaluation of information
visualizations
\cite{North:2006,Ellis:2006,Viegas:2006,Munzer:2009}.

@@

Here we suggest a new approach using recent research on visual inference \cite{buja:2009,wickham:2010}. In visual inference, data plots are considered to be test statistics, and these are compared with plots of data generated from a null hypothesis using a lineup. The null hypothesis underlying a lineup is that there is no real structure visible in the data plot, that any patterns seen are consistent with randomness, or from a known model. If the null hypothesis is true then the plot of the data will not be distinguishable from the plots of null data. A lineup is similar to the police lineup, and this is where the name comes from. The suspect (the plot of the data) is placed randomly in a field of plots of null data, and an impartial viewer is asked to identify the plot that is the most different. If the viewer identifies the suspect it lends statistical significance to the conclusion that the data is not consistent with the null data. Figure \ref{lineup} has an example of a lineup. Overlaid density plots are used to display the distribution of two groups (blue, red) of data. We are interested to know whether the data has a significant difference between the centers of the two groups. Plot number $5+2^3$ contains the data. Did you pick that one? In this plot the blue group has a flat density whereas the density of red group is more peaked and concentrated to the left of the blue.

\begin{figure}[htbp]
   \centering
   \includegraphics[height=\linewidth]{density_1_45_15_20_13.png} 
   \caption{Example lineup using density plots: one plot of data
     embedded with nineteen plots of null data. Which plot is
     most different from the others? (See the text for the answer.)}
   \label{lineup}
\end{figure}

The previous work \cite{buja:2009,wickham:2010} describes the use of lineups for evaluating discoveries made in exploratory analysis. Here, we describe the use of lineups for evaluating plot design for communicating findings visually. Different lineups are created, using the same data, same null plots and positions within the lineup, but different plot designs. Comparing how long viewers take and how accurately they report the feature of interest will assess which design is better for the task. To investigate the benefit of lineups for plot design evaluation we have conducted an experiment that also contains simulated data, where the data is constructed to match and expand on the features of interest in the data. For example, in a simple situation, where the discovered feature of interest is a difference between the centers of two distributions, simulated data would be constructed roughly matching the distribution of the data that varies the distance between centers, and also spread and sample size. Lineups of these simulated data (embedded with simulated null data plots) are generated and evaluated along with the lineups of data. The simulation study provides a backdrop for the real data, which enables the broad applicability to be studied and a gauge for the strength of the pattern in the data.

@@

Lineups provide a way to statistically quantify the significance of the finding \cite{majumder:2011}: think of the lineup as a set of $m$ plots, one of which is the data (in Figure \ref{lineup}, $m=20$).  The probability that an observer picks the data plot just by chance, i.e. when it is really *not* different from the other null plots, is $1/m$.  If an observer is able to identify the data plot from the lineup, we reject the null hypothesis.  This sets our Type I error rate, $\alpha$, at a level of $1/m$.

@@

The power of lineups is used in the work presented here to evaluate the effectiveness of different plot designs. This paper describes two examples where we had real data analysis problems and decisions to make in order to communicate the results. Section \ref{designs} explains the process of comparing designs. Section \ref{setup} describes two data analysis problems, the experiments conducted to evaluate the plot designs, and presents findings. We conclude in Section \ref{conclusions}  and give suggestions for future use.

@@

While this allows a direct comparison of the designs, we cannot adjust for the individuals' perceptual abilities.
In the case that we have multiple responses from each person (i.e. data is collected on several {\it different} lineup tasks), we can estimate their perceptual ability and correct power differences between competing designs accordingly, e.g. by modeling power using  a subject-specific random intercept in a generalized linear model.

### 2012-06-27 review/setup.tex
revision

% !TEX root = comparison.tex
\section{Experimental Setup}~\label{setup}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{website} 
   \caption{Screenshot of the website for the second study. }
   \label{fig:screen}
\end{figure}

Each participant is shown ten lineups and asked to identify the plot that is the most different from the other plots. In the second study the `difference' is further specified as `In which plot is the blue group the furthest to the right?". 
 In both studies, though, the question is placed prominently above each lineup as a reminder of what participants are asked to look for.  Additionally, participants are asked for a reason for their choice. They can select from a list of choices particular to the lineup design, why they chose this particular plot as well as how confident they are on a scale of one to five. Personal information on age group, gender, and education is collected on a voluntary basis.  The amount of time it takes the individual to answer is also recorded.  This  allows us to assess how design choices affect decoding \cite{cleveland:1994} in terms of accuracy and speed.

>>>>

% !TEX root = comparison.tex
\section{Experiments}~\label{setup}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{website} 
\vspace{-.2in}
   \caption{Screenshot of the website for the second study. }
   \label{fig:screen}
\end{figure}

Each participant is shown ten lineups and asked to identify the plot that is the most different from the other plots. In the second study the `difference' is further specified as `In which plot is the blue group the furthest to the right?". 
 In both studies, though, the question is placed prominently above each lineup as a reminder of what participants are asked to look for.  Additionally, participants are asked for a reason for their choice. They can select from a list of choices particular to the lineup design, why they chose this particular plot as well as how confident they are on a scale of one to five, where higher values indicate higher confidence. Personal information on age group, gender, and education is collected on a voluntary basis.  The amount of time it takes the individual to answer is also recorded.  This  allows us to assess how design choices affect decoding \cite{cleveland:1994} in terms of accuracy and speed.
